---
title: "Yambol Data Preparation"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Data Loading and Preparation

This script prepares the data for follow up analysis. It helps you load the FAIMS data from 2009-2018 and 2022 on burial mounds from Yambol as well as associated Yambol region vectors and elevation rasters. It uses the latter for mound data enrichment.
The needed inputs are:

- raw data on mounds (provided)
- Turkish and SE Bulgarian Aster images (not provided - LINK needed)


Its outputs are:

- an aggregate dataset of FAIMS features with attributes such as ASTER elevation, topographic prominence, roughness etc. 
- mosaiced raster that fits the confines of the Yambol Province
- mosaiced raster that crosses the Yambol-Turkish border 

## Setup

```{r libraries}
library(tidyverse)
library(sf)
library(raster)
```

## Load 2009-2018 data
This data is a bit special as it was processed in 2020 (see MoundMerging2020 repo in Github)
```{r mounds18}
# Environmental-extraction-enriched output from 2020 Mound Merging repo > 07_EnrichSpatial.R
mn09_18 <- read_csv2("../data/enriched_all.csv")
names(mn09_18)
mn09_18 %>% 
  group_by(Type) %>% 
  count()


# If you need raw data, see the raw/ folder
# mn17 <- read_csv("../data/2017ElenovoAll.csv")
# mn18 <- read_csv("../data/2018Bolyarovo.csv")
# mn10 <- read_csv("../data/2010-LGV-Bara-enrichedOR-20200619.csv")
# names(mn10)
# names(mn17)
```
## Load 2022 data

2022 data was freshly collected via FAIMS app in September 2022.

I used 2022_Elhovo_ORcleaningscript to clean up data from the Elhovo 2022 Burial module first (https://docs.google.com/document/d/1Xcq5yuQOrQpVOkGF2fp3ZeK4ynyaTVa3sZZ41NI7ZF8/edit) and deposit the exported spreadsheets in the data/ folder

```{r load-mnd22-data}
# Load data
mounds22_1 <- read_csv("../data/20220921BurialMoundOne-csv.csv")
mounds22_2 <- read_csv("../data/20220927BurialMoundTwo-csv.csv")

mounds22 <- rbind(mounds22_1,mounds22_2)
table(mounds22$TypeClean)
names(mounds22)

# wrangle to 20 columns
mnd22 <- mounds22 %>%
  rename(TRAP=MoundID, Type=TypeClean, LU_Around = LanduseAroundMound, LU_Top = LanduseOnTopOfMound) %>% 
  dplyr::select(TRAP, Source, Type, LU_Around, LU_Top, DiameterMax, HeightMax, Condition, PrincipalSourceOfImpact, Date, Northing, Easting, geospatialcolumn)
```

## Spatialize

```{r make-verified-spatial}
# check for missing coordinates
which(is.na(mn09_18$geometry))

# convert to a simple feature
mn18 <- st_as_sf(mn09_18, coords = c("X", "Y"), crs = 32635)

# sanity check
plot(mn18$geometry)

```

now, add 2022 data
```{r 22-spatial}
# check for missing coordinates
which(is.na(mnd22$geospatialcolumn))

# convert to a simple feature
mn22 <- st_as_sf(mnd22, coords = c("Easting", "Northing"), crs = 32635)

# sanity check
plot(mn18$geometry, col = 'darkred');plot(mn22$geometry, col = 'red', add =T)

rm(list = c("mounds22_1","mounds22_2", "mounds22", "mnd"))

```


## Yambol environmental and topographic data


### Load Vector data
```{r load-Y-vector}
# Load vectors
bg_border <- st_read("../data/Bulgaria_Border.shp")
Y_towns <- st_read("../data/Modern_settlements_Yambol_TRAP.shp")
#towns <- st_read("~/../Documents/Professional/Projects/MQNS/GIS/Vectors/Yamurbanarea.shp") # towns within 15km buffer of Yambol boundary exported from JICA
Y_region <- st_read("../data/YamRegion.shp")

```

### Mosaic and project component rasters
Wrangle rasters and mosaic them, then project to 32635.
Load vectors and check they are in 32635

```{r mosaic-rasters,eval =FALSE}
# Load ASTER DEM raster, provided by JICA (2005?), unprojected! (projected files do not align)

Yam <- raster("C:/Users/adela/Desktop/TRAP_Oxbow/YAM/ASTGTM_N42E026/ASTGTM_N42E026_dem.tif")
Tur <- raster("C:/Users/adela/Desktop/TRAP_Oxbow/YAM/ASTGTM_N41E026/ASTGTM_N41E026_dem.tif")

Yam # projected WGS84 0.00027 resolution, 3601x3601 cells
Tur # projected WGS84 0.00027 resolution, 3601x3601 cells

# Mosaic the two tiles to one covers the BG-TUR boundary 
library(raster)
library(rgdal)

YT_elev <- mosaic(Yam, Tur, fun = mean, tolerance = 1)
writeRaster(Y_elev, file="../output_data/Yelev.tif", format="GTiff", overwrite = TRUE)

# Project to 32635 so it's consistent with vector data
YT_elev32635 <- projectRaster(YT_elev, crs = 32635)

writeRaster(Y_Lelev32635, file= "../output_data/YT_elev32635.tif", format="GTiff", overwrite = TRUE)

# Crop to Yambol region boundary
Y_elev <- crop(YT_elev32635, Y_region)
Y_elev <- mask(Y_elev, Y_region)
writeRaster(Y_elev, "../output_data/Yelev32635.tif", format="GTiff", overwrite = TRUE)

rm(Yam)
rm(Tur)
rm(Y_elev)
rm(YT_elev)
rm(YT_Lelev32635)
```


### Load Raster data
```{r load-elev}
library(raster)

# Load projected raster cropped to Yambol region boundary 

Y_elev32635 <- raster("../output_data/Yelev32635.tif")
plot(Y_elev32635)


# Load large mosaiced raster ( "../output_data/YT_elev32635.tif" for both Turkey & SE Bulgaria) because you have mounds outside Yambol region, rename and crop it to a buffer around Yambol 

elev <- raster("../output_data/YT_elev32635.tif")
plot(elev)
# elev <- crop(Y_Lelev32635, st_buffer(Y_region, 5000))
# elev <- mask(elev, st_buffer(Y_region, 5000))
# plot(elev)


```

## Plot everything
```{r plot-all}
plot(elev)
#plot(towns$geometry, col = 'green', add =T);
# plot(Y_towns$geometry, col = 'red', add =T)
plot(bg_border$geometry, add = T)
plot(mn18$geometry, add = T)
plot(mn22$geometry, add =T)

```


## Enrich 2022 data
2018 features have already been enriched during their processing in 2020. 2022 mounds need this basic enrichment. Both datasets will need TPI, TRI and other indeces extracted from the raster data.


### Distance to BG border

```{r}
# Distance from points to BG boundary
mn22$distBG <-  st_distance(mn22,bg_border)
mn22$distBG[1:3,]

mn18$distBG <-  st_distance(mn18,bg_border)
mn18$distBG[1:3,]

```


### Distance from mound points to nearest settlement

This is a two-step process: 

1. selecting closest town via `st_nearest()` and 
 `nearest <-  try(st_nearest_feature(mn22, towns))`
 try()` is a wrapper to run an expression that might fail and allow recovery
2. getting distance with `st_distance()`
 `distTown <- st_distance(mn22, towns[st_nearest_feature(mn22, towns),], by_element = TRUE)`
 
```{r distance}
# Distance to nearest (as the crow flies) village
mn22$distTown <- st_distance(mn22, towns[st_nearest_feature(mn22, towns),], by_element = TRUE)
head(mn22,3)
mn18$distTown <- st_distance(mn18, towns[st_nearest_feature(mn18, towns),], by_element = TRUE)
head(mn18,3)
```



### Sample elevation, aspect and slope at mound coordinates for 2022 data

```{r elev-aspect-slope-22}
# Prerequisite to sampling - a dataframe with coordinates alone
mound_coordinates <- data.frame(st_coordinates(mn22)) # don't add TRAP yet

# Sample elevations at mound locations
mn22$elevAster <- raster::extract(elev, mound_coordinates)

# Create slope and aspect raster (can add 'flowdir' also, if needed)
Y_aspslope <- raster::terrain(elev, opt = c('slope','aspect'), neighbors = 8, unit= 'degrees')
Y_flow <-  raster::terrain(elev, opt ='flowdir', neighbors = 8) #flowdir returns the 'flow direction' (of water), i.e. the direction of the greatest drop in elevation (or the smallest rise if all neighbors are higher). They are encoded as powers of 2 (0 to 7). The cell to the right of the focal cell 'x' is 1, the one below that is 2, and so on:
Y_TPI <- raster::terrain(elev, opt ='TPI', neighbors = 8)
Y_TRI <- raster::terrain(elev, opt ='TRI', neighbors = 8)
Y_rough <- raster::terrain(elev, opt ='roughness', neighbors = 8)

# If you want to see the output
# plot(elev)
# plot(Y_aspslope)
# plot(Y_flow)
# plot(Y_TRI)

# ..with the mounds
# plot(mn22$geometry);plot(Y_flow, add =T); plot(mn22$geometry, add= TRUE)


# Extract values
mn22$slopeAster <-  raster::extract(Y_aspslope$slope, mound_coordinates)
mn22$aspectAster <-  raster::extract(Y_aspslope$aspect, mound_coordinates)
mn22$TRI <-  raster::extract(Y_TRI$tri, mound_coordinates)
mn22$TPI <-  raster::extract(Y_TPI$tpi, mound_coordinates)
mn22$rough <-  raster::extract(Y_rough$roughness, mound_coordinates)

```

### Calculate prominence 
```{r prominence}
library(FSA) # we need an additional library for the perc() function
?perc()      # check the fce is here
mn22$prom250mbuff <- raster::extract(elev,# raster containing elevation data
                        mound_coordinates, # centroids of mounds
                        buffer = 250, # actual buffer size in crs units, in this case 250m  or ca 7x7 cells around kernel
                        fun = function(x){perc(x,x[length(x)/2],"lt", na.rm = FALSE, digits = 2)})

summary(mn22[,13:19]) 

```

## Enrich 2018 data

it was already done in 2020, but let's do it again for consistency's sake

```{r sample-elev-aspect-slope-18}
# Prerequisite to sampling - a dataframe with coordinates alone
mound_coordinates <- data.frame(st_coordinates(mn18)) # don't add TRAP yet


# Sample elevations at mound locations
mn18$elevAster <- raster::extract(elev, mound_coordinates)

# Sample slope and aspect values
mn18$slopeAster <-  raster::extract(Y_aspslope$slope, mound_coordinates)
mn18$aspectAster <-  raster::extract(Y_aspslope$aspect, mound_coordinates)
mn18$TRI <-  raster::extract(Y_TRI$tri, mound_coordinates)
mn18$TPI <-  raster::extract(Y_TPI$tpi, mound_coordinates)
mn18$rough <-  raster::extract(Y_rough$roughness, mound_coordinates)

# Calculate prominence 

library(FSA) # we need an additional library for the perc() function
?perc()      # check the fce is here
mn18$prom250mbuff <- raster::extract(elev,# raster containing elevation data
                        mound_coordinates, # centroids of mounds
                        buffer = 250, # actual buffer size in crs units, in this case 250m  or ca 22x22 cells around kernel
                        fun = function(x){perc(x,x[length(x)/2],"lt", na.rm = FALSE, digits = 2)})


summary(mn18[,16:24])

```


## Combine enriched datasets

```{r combine-18-22}

names(mn22) # 18
names(mn18) # TopoID, Note, ...1, prom250mbuff
str(mn22)
str(mn18)

# Streamline columns before merging
mn18 <- mn18 %>% 
  dplyr::select(TRAP, Source, Type, LU_Around, LU_Top, DiameterMax, HeightMax, Condition, PrincipalSourceOfImpact, geometry, distBG, distTown, elevAster, slopeAster, aspectAster, prom250mbuff, TPI, TRI, rough)

mn22 <- mn22 %>% 
  dplyr::select(TRAP, Source, Type, LU_Around, LU_Top, DiameterMax, HeightMax, Condition, PrincipalSourceOfImpact, geometry, distBG, distTown, elevAster, slopeAster, aspectAster, prom250mbuff, TPI, TRI, rough)

# Merge datasets
mnds <- rbind(mn18,mn22)

# Sanity check to see if all looks good
plot(mnds$geometry)
str(mnds)
summary(mnds[,11:19])

```

## Regenerate distances as they got truncated

```{r mnds-distances}
# Distance from points to BG boundary
mnds$distBG <-  st_distance(mnds,bg_border)
# Distance from mound points to nearest settlement
mnds$distTown <- st_distance(mnds, towns[st_nearest_feature(mnds, towns),], by_element = TRUE)
summary(mnds[,11:19])
```




```{r}
library(mapview)
mapview(mnds %>% filter(Type == "Burial Mound"), zcol = "distBG")
```
## Export enriched dataset

Export the interim enriched set of all features (inside and outside of Yambol) collected during 2009-2022 seasons in the Yambol region. These will need to be filtered for actual mounds as there are settlement mounds, bunkers and lots of other stuff in there. They also need streamlining in terms of Condition, Height, etc. 
```{r interim-export, eval = FALSE}
write_rds(mnds, "../output_data/mnds_aggr.rds")
read_rds("../output_data/mnds_aggr.rds")
```

## Export rasters as a brick
```{r}
Y <- brick(Y_flow, Y_rough, Y_TPI, Y_TRI)

# Write to file

writeRaster(Y, filename = "../output_data/YflowroughTRI.tif", format="GTiff", overwrite = TRUE)
writeRaster(Y_aspslope, filename = "../output_data/Yaspslope.tif", format="GTiff", overwrite = TRUE)
rm(list = c("Y", "Y_rough", "Y_TPI","Y_TRI", "Y_flow"))
rm(Y_aspslope)
```



## Add data necessary for AKB NEEDS WORK
### Distance from each mound to its "zemlishte"

AKB asks for distance and azimuth to the nearest cadastral village (in "zemlishte"). The tricky part is that the cadastral villages need not be the geographically nearest places, and so we need to know which zemlishte territory is each mound in so we can calculate the distance between it and all the mounds in the village cadastre.

The information about which village cadastre does a mound belong to can be derived from an intersection of the mounds and zemlishte polygons. Here we use a shortcut: spatial points output from a spatial query in ArcGIS (ElhStr_features.shp), which contains the TRAP ID of each mound and the Name and number of the zemlishte village (Name_en and EKATTE)
We have the zemlishte village point data with Name_en and EKATTE in ElhStr_towns shapefile

```{r load-zemlishe}
# Zemlishte information sit in ElhStr_towns shapefile
village <- read_sf("../data/ElhStr_towns.shp")
data22 <- read_sf("../data/ElhStr_features.shp")
```

### Wrangle the features and nearest zemlishte 
```{r wrangle-zemlishte}
# Subset the features 2022 data to only TRAP numbers and the name of the  zemlishte-defining village
features22 <- data22 %>%
  dplyr::select(TRAP,EKATTE, Name_en) #EKATTE and Name_en are the number and name of the zemlishte that we are interested in

# Create a simple feature of village points ordered by the TRAP numbers from the features 22 dataset,   
villages22 <- features22 %>% 
  st_drop_geometry() %>%  # dropping feature geometry
  left_join(village, by = "EKATTE") %>%  #joining village geometry to TRAP numbers
  dplyr::select(-Descr_bg, -Name_bg, -Suffix_bg, -Suffix_en)
villages22
```

### Calculate distances

Once we have the starting and ending points, the distances are trivial to calculate between mounds and zemlishte using the st_distance() function, utilizing by_element argument, because I am only interested in pairwise distances 
```{r distance-zemlishte}
features22$distToTown <- st_distance(features22$geometry, villages22$geometry, by_element = TRUE)
ls('package:sf')
```

### Calculate azimuth

Azimuth can be tricky as it is an angular measurement which is often produced only in geographic space and here we have geometric, planar space. I use the `nngeo` package of Michael Dorman as it calculates "planar" /geometric azimuth (https://michaeldorman.github.io/nngeo/index.html).
sf library works in geometric /planar space and so 
The other `geosphere::azimuth` which calculates geographic azimuth in `latlong` coordinates.(https://stackoverflow.com/questions/51030060/in-geosphere-package-in-r-why-arent-bearings-in-0-360-degrees) 


```{r azimuth}
install.packages("remotes")
remotes::install_github("michaeldorman/nngeo")
features22$azimuth <- nngeo::st_azimuth(features22$geometry, villages22$geometry)
```

### Export for Toshko
```{r export}
# All 479 features from the Elhovo and Straldzha municipality (2010-2022)
write.csv(features22 %>% st_drop_geometry(), "../output_data/features22forToshko.csv")

# Only 238 features collected in 2022 in the Elhovo and Straldzha municipality
library(mapview)
features22 %>% 
  filter(TRAP %in%mnd22$TRAP) %>% 
  mapview()

# Only 311 features collected in 2010-2022 in the Elhovo municipality
mun <- read_sf("../data/Yam_Municipalities.shp")

library(sf)
features22$geometry %>% 
  sf::st_intersection(mun[4,]) %>% 
  mapview()
```
