---
title: "Wrangle and Visualise Yambol Mounds"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Wrangling Mound data out of FAIMS data

The purpose of this script is to clean up and filter the total aggregate of FAIMS data originating from the Yambol region to mostly-mound points from within the regional boundary, whose attributes are analysis-ready.

## Setup
```{r libraries}
library(tidyverse)
library(sf)
```


## Load data
```{r -load-aggr-data}
features <- read_rds("../output_data/features_dd_later_27Dec.rds")
str(features)
#colnames(features)
#st_write(features, "../output_data/features_aggr_27Dec.shp")
```

## Validation I: Check how many of the features are actually mounds

```{r type-tally}
# Filter mounds by type
features %>% 
  group_by(Type) %>% 
  tally()
```
```{r type-check}
# Verify that "Other" category is not legacy-data verificaiton, because those should be mounds/extinct mounds if the symbol was a brown sunburst

features %>% 
  filter(Type == "Other") %>% 
  group_by(Source) %>% 
  tally()

```
In the tally of "Other" types, there are 34 Legacy verification features and 41 Survey features. While the latter are expected, the 34 LV features require follow up. TO BE DONE!! 

## Validation II: Check for duplicates (similar attributes)
Spatial duplication is addressed in 00b_SpatialDeduplication.Rmd
```{r duplicate-check}
features$TRAP[duplicated(features$TRAP)] 
```

## Validation III: Clean up condition, height and other attribute

```{r clean-up}
features <- features %>%
  #st_drop_geometry() %>%  
  dplyr::mutate(Condition = str_extract(Condition, "\\d")) %>%
  dplyr::mutate(Condition = case_when(Condition == 0 ~ "NA",
                               Condition == 6 ~ "5",
                               Condition != 0 ~ Condition)) %>% 
  dplyr::mutate(Condition = as.factor(Condition)) %>% 
  dplyr::mutate(Type= gsub("\\?","",Type)) %>% 
  dplyr::mutate(HeightMax = as.numeric(HeightMax)) %>% 
  dplyr::mutate(DiameterMax = as.numeric(DiameterMax)) 

#levels(features$Condition) = c(1,2,3,4,5, NA)

# See if you need to eliminate uncertainty from Height
# unique(features$HeightMax) # are there any ? in the data values?
# features %>%  # if the answer is yes, you can filter them out.
#   mutate(HeightMax = gsub("\\?", "",HeightMax)) %>%
#   mutate(HeightMax = as.numeric(HeightMax))
features %>% 
  group_by(Type) %>% 
  tally()
```
### Quick visual

```{r}
library(mapview)
mapview(features, zcol = "Type")
```

### Subset to mounds only

```{r filter-mounds}
mnds <- features %>% 
  dplyr::filter(Type == "Burial Mound" | Type == "Extinct Burial Mound" | Type == "Uncertain Mound")

str(mnds)
```

## Clip data by Yambol region
```{r yambol-mnds}
Y_region <- st_read("../data/YamRegion.shp")
Yam_mnds <- st_intersection(mnds, Y_region$geometry)
plot(Yam_mnds$geometry);
plot(Yam_mnds["Condition"])
```

## Export the Yambol mound dataset

now that all looks reasonably well, let's export the data. These are features inside the Yambol Province.

```{r export-mnds}
write_rds(Yam_mnds, "../output_data/Yam_mnds.rds")
#st_write(Yam_mnds,"../output_data/Yam_mnds.shp", append = FALSE )

```


## Feature types
Yam_mnds is still a sf feature.
```{r tally-mnds-types}
Yam_mnds %>% 
  group_by(Type) %>% 
  tally()

```

## Filter mounds, including extinct ones
```{r filter-extinct}
Yam_mnds <- Yam_mnds %>% 
  #st_drop_geometry() %>% 
  dplyr::filter(Type == "Burial Mound" | Type == "Extinct Burial Mound" | Type == "Uncertain Mound")

str(Yam_mnds)

```

## Mound sizes
```{r mound-size}
Yam_mnds %>%
  group_by(Type) %>%
  summarize(Max = max(HeightMax, na.rm = T), 
            Mean = mean(HeightMax, na.rm = T),
            Median = median(HeightMax, na.rm = T))

```
## Mound size boxplots
```{r size-boxplot}

Yam_mnds %>%
    ggplot(aes(x = Type, y = HeightMax)) +
    geom_violin()+
    geom_jitter(alpha = 0.3,
    		color = "tomato",
    		width = 0.2,
    		height = 0.2) +
   theme_bw() +
  labs(x = "Mound type",
       y = "Maximum recorded height",
       title = "Mound size by feature type")
```

```{r size-histogram}
Yam_mnds %>% 
  ggplot(aes(x = HeightMax)) +
  geom_histogram(aes(fill = Type)) +
  theme_bw() +
  labs(x = "Mound height (m)",
       y = "Feature height recorded during visit",
       title = "Mounds in the Yambol Region")


```


## Mound condition

```{r condition-plot}
plot(Condition ~ HeightMax, Yam_mnds)
```
## Plot Condition in Space

```{r map-plot}
plot(Yam_mnds["Condition"])
plot(Yam_mnds["elevAster"])
```

## Static map
```{r get-yambol-elev}
library(raster)
library(sf)

Y_elev <- raster("../output_data/large/Yelev32635.tif")
st_crs(Y_elev)==st_crs(Yam_mnds)

plot(Y_elev)
plot(Yam_mnds$geometry,cex = sqrt(Yam_mnds$HeightMax), add =T)
```

## Interactive maps

### Mounds by height

```{r mapview-height}
library(mapview)
class(Yam_mnds)
mapview(Yam_mnds, zcol="HeightMax")
```

### Mounds and their condition
```{r condition-map}
library(mapview)
mapview(Yam_mnds, zcol="Condition") + mapview(Y_elev)
```

