---
title: "Wrangle Yambol Mounds - early version"
author: "Adela Sobotkova"
date: "updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Wrangling Mound data out of FAIMS data

The purpose of this script is to clean up and filter the total aggregate of FAIMS data originating from the Yambol region to mostly-mound points from within the regional boundary, whose attributes are analysis-ready and whose spatial points are deduplicated.

This script uses the early records in case of the 18 spatial duplicates. For comparison on the stats, please use with script 01a

## Setup
```{r libraries}
library(tidyverse)
library(sf)
library(mapview)
```


## Load data
```{r -load-aggr-data}
features <- readRDS("../../MoundMerging2023/output_data/features_dd_later.rds") # 2023

str(features)

```

## Validation I: Check how many of the features are actually mounds

```{r type-tally}
# Filter mounds by type
features %>% 
  group_by(Type) %>% 
  tally()
```
```{r type-check}
# Verify that "Other" category is not legacy-data verificaiton, because those should be mounds/extinct mounds if the symbol was a brown sunburst

features %>% 
  filter(Type == "Other") %>% 
  group_by(Source) %>% 
  tally()

```
In the tally of "Other" types, there are 34 Legacy verification features and 41 Survey features. While the latter are expected, the 34 LgV features required follow up on 27 Dec 2022. Inspection showed that many of the LGV features originate not from sunbursts but other map markers, such as raysed squares and triangles, which were demonstrated to lead to mounds less reliably. 

### what is "other"?
```{r}
features %>% 
  filter(Type == "Other" & Source == "Legacy verification") %>% 
  group_by(PrincipalSourceOfImpact) %>% 
  tally()
```


## Validation II: Check for duplicates (similar attributes)
Spatial duplication is addressed in 00b_SpatialDeduplication.Rmd
```{r duplicate-check}
features$TRAP[duplicated(features$TRAP)] 
```

## Validation III: Clean up condition, height and other attribute

```{r clean-up}
unique(features$Condition)

features <- features %>%
  #st_drop_geometry() %>%  
  dplyr::mutate(Condition = str_extract(Condition, "\\d")) %>%
  dplyr::mutate(Condition = case_when(Condition == 0 ~ "NA",
                               Condition == 6 ~ "5",
                               Condition != 0 ~ Condition)) %>% 
  dplyr::mutate(Condition = as.factor(Condition)) %>% 
  dplyr::mutate(TypeCertainty = case_when(
    grepl("\\?",Type) ~ "Uncertain",
    !grepl("\\?",Type) ~ "Certain")) %>% 
  dplyr::mutate(Type= gsub("\\?","",Type)) %>% 
  dplyr::mutate(HeightMax = as.numeric(HeightMax)) %>% 
  dplyr::mutate(DiameterMax = as.numeric(DiameterMax)) 

#levels(features$Condition) = c(1,2,3,4,5, NA)

# See if you need to eliminate uncertainty from Height
# unique(features$HeightMax) # are there any ? in the data values?
# features %>%  # if the answer is yes, you can filter them out.
#   mutate(HeightMax = gsub("\\?", "",HeightMax)) %>%
#   mutate(HeightMax = as.numeric(HeightMax))
features %>% 
  group_by(Type) %>% 
  tally()
```

## Clip Features by Yambol boundary

```{r}
Y_region <- st_read("../data/YamRegion.shp")
Y_features <- st_intersection(features, Y_region$geometry)
Y_features
```
Within Yambol we documented 1243 features. 

## Source

### all features
```{r source-all-features}
features %>% 
  group_by(Source) %>% 
  tally()
```

In terms of feature source, 1046 map features from the entire visited area (Yambol and surrounding) were visited, and 397 moundlike features were discovered along the route. 

Let's see what the stats are for Yambol Province.

### Source of Features within Yambol
Were the features found thanks to topo maps or via survey?

```{r source-Yam}
Y_features %>% 
  group_by(Source) %>% 
  tally()
```

In Yambol Province, we have documented 1243 features, 847 originated from map symbol and the rest were encountered en route by means survey (373), local informants and serendipity.


```{r type-yambol}
Y_features %>% 
  group_by(Type, TypeCertainty) %>% 
  tally()

Y_features %>% 
  filter(grepl("Mound", Type))
```

## Explore Uncertain Features - move to MoundMerging2023 once finished!
This is to relabel features if they do not conform to the premises of the project (ie. are in map, but marked not as extinct mounds)
```{r}
Y_features %>% 
  filter(TypeCertainty == "Uncertain",  Source == "Legacy verification") %>% 
  filter(grepl("Mound|Uncertain", Type)) %>%
  #group_by(Type) %>% tally
  st_drop_geometry()
  
# Export uncertain Burial Mounds from LGV as shp to check in GEarth and images
Y_features %>% 
  filter(TypeCertainty == "Uncertain",  Source == "Legacy verification") %>% 
  filter(grepl("Mound|Uncertain", Type)) %>%
  select(-TopoID) %>% 
  st_write("../output_data/LGVMoundsUncertain.shp", append=F)

```
9516 : photo - typical feature of BG fields, a small-mound sized cluster of stones hiding in a scrub in a slight terrain wave,  could be natural geology, a field stone cluster, or an extinct mound
8337 : terrain wave, possible mound, in GE satellite image a round light soil mark 80m to SW from GPS point 
8388 : terrain wave near a serious mound 8387, nothing in GE satellite image now, but soilmark in 2009
9645 : water station / extinct mound? next to a road
8427 : moundlike feature (round, conical pile) with concrete construction
9838 : ploughed up mound (definitely)l light soil marks in GE >> CHANGE UNCERTAIN TO CERTAIN
9847 : pile in a forest; no visibility; should be UNCERTAIN
9852 : pretty clear scrub covered mound! l    ozenge in GE >> CHANGE UNCERTAIN TO CERTAIN
8764 : very uncertain lozenge between two mounds: natural or cultural >> should be SURVEY (instead of 8763) AND UNCERTAIN EXTINCT mound
8766 : massive mound, why the doubt? >> CHANGE UNCERTAIN TO CERTAIN
8770 : two concentric circles, negative and positive: strange, something else?
8772 : smallish mound like protrusion > marked as extinct and should stay so
8775 : hardly anything >> maybe eliminate or keep as extinct
8648 : disturbed area >> should perhaps be eliminated
9883 : flat nothing in photo or GE; lots of pottery >> CHANGE TO SCATTER
9910 : very low feature in photo, light soilmark in GE, next to a solid 9909 >> CHANGE TO EXTINCT
9911 : looks like nothing, check Topo map, might need flagging at EXTINCT or deletion



### Quick visual

```{r}
library(mapview)
mapview(Y_features, zcol = "Type")
```
### Export Yambol Features
```{r}
st_write(Y_features, "../output_data/features_dd_Yam.geojson")

Y_features %>% 
  mutate(latitude = st_coordinates(.)[,2],
               longitude = st_coordinates(.)[,1]) %>% 
  st_drop_geometry() %>% glimpse()

```

### Check how many LGV and survey features have 0 height

```{r}
Y_features %>% 
  filter(grepl("Mound|Uncertain", Type)) %>% 
  filter(HeightMax < 0.2) %>% 
  group_by(Source, Type) %>% 
  tally()

Y_features %>% 
  st_drop_geometry() %>% 
  filter(Type == "Burial Mound") %>% 
  filter(HeightMax <0.2)
```


### Subset to mounds only

```{r filter-mounds}
mnds <- features %>% 
  dplyr::filter(Type == "Burial Mound" | Type == "Extinct Burial Mound")

str(mnds)
```

## Clip Mounds by Yambol region
```{r yambol-mnds}
Y_region <- st_read("../data/YamRegion.shp")
Yam_mnds <- st_intersection(mnds, Y_region$geometry)
plot(Yam_mnds$geometry);
plot(Yam_mnds["Condition"])
```

## Export the Yambol mound dataset

now that all looks reasonably well, let's export the data. These are features inside the Yambol Province.

```{r export-mnds}
write_rds(Yam_mnds, "../output_data/Yam_dd_early_mnds.rds")
#st_write(Yam_mnds,"../output_data/Yam_mnds.shp", append = FALSE )

```



## AKB mound joining
```{r}
# load the 1058 records of AKB numbers
AKB <- read_csv("../data/AKBmounds.csv")

# clean up the AKB Number column to discard type notes
AKB$`AKB №` <- as.numeric(AKB$`AKB №`)

# this coerces 349 records to NA
sum(is.na(AKB$`AKB №`))

# leaving 709 AKB records
AKB %>% 
  filter(`AKB №`> 0) %>% 
  filter(!is.na(Excavated)) # 12 mounds have been excavated
```
## Clean AKB up
```{r}
AKB <- AKB %>% 
  rename(SeasonNotes = Season, AKB = `AKB №`, TRAP = `TRAP №`)
```

## Join up
```{r}
Y_features <- Y_features %>% 
  left_join(AKB, by = c('TRAP'))

colnames(Y_features)

```
## CHeck how many are excavated
```{r}
Y_features %>% 
  filter(AKB > 0) %>% 
  filter(!is.na(Excavated)) # 11

Y_features %>% 
  filter(AKB > 0)  # 698 features have AKB >> duplicates must be here

```

