---
title: "Locational choices of mound builders seen through frequency analysis"
author: "Adela Sobotkova"
date: "2022-11-10"
output: html_document
---

This script attempts to approach the locational preferences of burial mound builders in Yambol from a different perspective, one that compares the frequency of mounds in different elevation bands with the frequency of this particular band in the Yambol region landscape. My hope is that this approach will be slightly more sensitive to the actual occurrence of mounds in the landscape than the Monte Carlo bootstrap in 02-03 scripts


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Extract regional background data from Yambol Province raster

We extract elevation, and create and extract slope, aspect and flowdir from the Aster cropped to the Yambol province boundaries, so as to have baseline/background data for frequency analysis
```{r region-extract}
# extract elevation
plot(Y_elev)
data_Y<-as.data.frame(Y_elev) 

# generate slope, aspect and flowdir data
Y_aspslope <- raster::terrain(Y_elev, opt = c('slope','aspect'), neighbors = 8, unit= 'degrees')
Y_flow <-  raster::terrain(Y_elev, opt ='flowdir', neighbors = 8)


# extract slope, aspect and flowdir into a table
data_Y <- cbind(data_Y, as.data.frame(Y_aspslope), as.data.frame(Y_flow))
## Extract data from Yambol Province raster
```



## Process background raster values
..into binned frequencies for frequency comparison. First, we explore the ranges and prepare the bins

```{r}
# explore the ranges
hist(data_Y$Yelev32635)
hist(data_Y$slope)
hist(data_Y$aspect)
hist(data_Y$flowdir)

```
Now , seeing what the min and max values are, we prepare te number sequences
```{r}
## Prepare number sequences for binning of actual raster values
elev_breaks <- c(0,100,200,300,400,500,600,700) # use for elevation
slope_bin <- c(0,2,4,6,8,10,20,60) #for slope values
aspect_bin <- c(0, 45, 135, 225, 315, 360)

```

Now for each data category, we create the bins using the `cut()` function and `dig.lab=number` of numbers we don't want abbreviated

## Bin Elevation, Slope, Aspect and Flow direction
```{r}
print("Calculating frequencies for ASTER elevation, slope and aspect data and binning them")
```

### Eliminate missing values
```{r}
# Check for missing data and eliminate (although it is not entirely necessary)
is.na(data_Y$Yelev32635)
e <- na.omit(data_Y$Yelev32635)

# Test bin creation with median value

tapply(e,cut(e,elev_breaks),median)
tapply(data_Y$slope,cut(data_Y$slope,breaks=slope_bin),median)
```
### Bin Elevation values
```{r}
elevbin <- cut(e,breaks = elev_breaks,dig.lab = 5) ## very important to make thousands appear not as scientific notations
df_yam_e <- data.frame(table(elevbin))   # calculate frequency via table() and convert to dataframe for printing
df_yam_e
```

### Bin Slope values
```{r}
(slopebin <- cut(data_Y$slope,breaks=slope_bin))
df_yam_sl <- data.frame(table(slopebin))
df_yam_sl
```

### Bin Aspect values
```{r}
aspectbin_yam <- cut(data_Y$aspect, breaks=aspect_bin)
df_yam_asp <- data.frame(table(aspectbin_yam))
df_yam_asp
```

### Bin Flowdir values
```{r}
flowbin_yam <- cut(data_Y$flowdir, breaks=aspect_bin)
df_yam_fl <- data.frame(table(aspectbin_yam))
df_yam_fl
```
## Bin Mound environmental variables
### ASTER Mound data 
```{r}
elevbins_mnd <- cut(Yam_mnds$elevAster,breaks=elev_breaks,dig.lab=5)
df_mnd_e <- data.frame(table(elevbins_mnd))
df_mnd_e

slopebin_mnd <- cut(Yam_mnds$slopeAster, breaks = slope_bin)
(df_mnd_sl <- data.frame(table(slopebin_mnd)))

aspectbin_mnd <- cut(Yam_mnds$aspectAster, breaks = aspect_bin)
(df_mnd_as <- data.frame(table(aspectbin_mnd)))
```

## Generate Aster frequences and output into a table
```{r combine-results}
print("Combining outputs into tables")
```

### Elevation

```{r}
Elev_freq <- cbind(df_yam_e,df_mnd_e[,2])
Elev_freq

colnames(Elev_freq) <-c("Elevation Bins","Yambol Region","Mounds")
head(Elev_freq)
```


### Slope

```{r}
Slope_freq <- cbind(df_yam_sl, df_mnd_sl[,2])
Slope_freq
colnames(Slope_freq) <-c("Slope Bins (degrees)","Yambol Region","Mounds")
head(Slope_freq)
```

### Flowdir

```{r eval=FALSE}
Flow_freq <- cbind(df_yam_fl, df_mnd_fl[,2])
Flow_freq
colnames(Aspect_freq) <-c("Aspect Bins (azimuth)","Yambol Region","Mounds")
head(Aspect_freq)

```


### Write Frequencies to file
```{r}
print("Writing results to file")

if(dir.exists("../output_data/")) {
   print("Output directory detected. Carrying on...")} else {
   dir.create("../output_data")
     }
  
write.csv(Elev_freq,"../output_data/YamMapMndElevationFrequency.csv")
write.csv(Slope_freq,"../output_data/YamMapMndSlopeFrequency.csv")
write.csv(Aspect_freq,"../output_data/YamMapMndAspectFrequency.csv")

print("All done with Frequency calculations. Proceeding onto Indeces")
```

Now that we have calculated the frequencies, let's see how they compare to one another.Are the mound locations dramatically different from the overall distribution of different natural aspects in the landscape?

## Indeces

###  Calculate Mound Location index in relief
An index is a number between 0-1 that marks the proportionality of mound locations in a given elevation/aspect or slope compared to the presence of this phenomenon in the landscape.

## Elevation 
```{r}
print("Creating an index of elevation frequencies for mounds")

ElevationResult <- Elev_freq %>% 
  mutate(AsterArea_km = `Yambol Region` *(0.027^2), 
         Yam_Percent =`Yambol Region`/sum(`Yambol Region`), 
         MoundsExpected =sum(Mounds)*Yam_Percent,
         MoundIndex = Mounds/MoundsExpected) %>% 
  mutate_at(4:7,round,2)

print("Printing ElevationResult to file.")
ElevationResult
write_csv(ElevationResult,"../output_data/ElevationIndeces.csv")


```
## Slope 
```{r}
print("Creating an index of slope frequencies for mounds")

SlopeResult <- Slope_freq %>% 
  mutate(AsterArea_km = `Yambol Region` *(0.027^2), 
         Yam_Percent =`Yambol Region`/sum(`Yambol Region`), 
         MoundsExpected =sum(Mounds)*Yam_Percent,
         MoundIndex = Mounds/MoundsExpected) %>% 
  mutate_at(4:7,round,2) 
  

print("Printing Slope Result to file.")
SlopeResult
write_csv(SlopeResult,"../output_data/SlopeIndeces.csv")

```

## Aspect 
```{r}
print("Creating an index of aspect frequencies for mounds")

Aspect_freq
Aspect <- c("North (315-45)" , "East (45-135)", "South (135-225)", "West (225-315)")

Aspect_final <- Aspect_freq %>%
  select(2:3) %>% 
  rbind(Aspect_freq[1,2:3] + Aspect_freq[5,2:3]) %>%   # adding row 1 and 4 as they are each half of North azimuth
  slice(c(2:4,6)) %>%  # drop rows 1 and 4 and replace by new row 6 for North
  cbind(Aspect) # add labels
Aspect_final
  
AspectResult <- Aspect_final %>% 
 mutate(AsterArea_km = `Yambol Region` *(0.027^2), 
         Yam_Percent =`Yambol Region`/sum(`Yambol Region`), 
         MoundsExpected =sum(Mounds)*Yam_Percent,
         MoundIndex = Mounds/MoundsExpected) %>% 
  mutate_at(5:7,round,2) 
 
AspectResult

```

```{r}
print("Printing Aspect Result to file.")
write_csv(AspectResult,"../output_data/AspectIndeces.csv")
print("Calculation completed; Check Output_data folder for results")
```


