---
title: "Visibility"
author: "Adela Sobotkova"
date: "2022-12-14"
output: html_document
---
## Purpose and prerequisites:
This script serves to elaborate the calculation of mutual intervisibility (clear line of sight) between various archaeological features, starting here with burial mounds in the Yambol Region and ASTER elevation raster (30m spatial resolution). 

The script basically works, calculating linestrings, intervisibility between Yam_mounds and generating maps

The proof of concept is done on BA mounds and to replicate it you might need: 
- to run 09 BA mounds to have the necessary libraries and digital objects

THe code should be expandable
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE)
library(sf)
library(raster)
library(tidyverse)
```

## Load 2009-2022 data
```{r load-mounds}

Y_elev <- raster("../output_data/large/Yelev32635.tif")
# Yambol mounds
Yam_mnds <- readRDS("../output_data/Yam_dd_mnds.rds") # 1073 features
Yam_mnds
Yam_mnds %>% 
  group_by(Type) %>% 
  tally()
```


# Intervisibility of Bronze Age mounds in Yambol
In this proof of concept we explore the visibility of BA mounds from other BA mounds, for simplicity's sake. 

BA mounds comprise surveyed and excavated mounds ('bronze') and excavated prior to survey ('extrasf') supplied by Toshko.

Excavated prior:
Drazhevo - Height: 2.7 m; Diameter: 37.0 m; Land: pasture on the rocky area
Straldzha - Height: 4.0 m; Diameter: 50.0 m; Land: wines and pasture
Popovo - Height: 6.8 m; Diameter: 48.0 m; Land: wines and pasture

8345 Mogila village - Height: 3.2 m; Diameter: 37.6 m; Land: pasture on the rocky area 


```{r BA-faims, include = FALSE}
# Filter out the BA mounds from the FAIMS data
bronze <- c(6009,8007,6010,8345,8346,8357, 8502, 8351, 9257)
BA <- Yam_mnds %>% 
  filter(TRAP %in% bronze)

# Add literature dimensions to mound 8345, which was excavated before survey 
BA$DiameterMax[5] <- 37.5
BA$HeightMax[5] <- 3.2

```

```{r BA-Todor, include = FALSE}
# Add the few mounds excavated before TRAP coverage >> these should be pulled out of the AOR dataset to get the Height, Diameter and other dimensions..
extras <- data.frame(place = c("Drazhevo", 
                           "Straldzha",
                           "Popovo"),
                 Easting=c(455442,479522,479391),
                 Northing=c(4710935,4715730,4668279),
                 HeightMax=c(2.7,4.0,6.8),
                 DiameterMax = c(37,50,48),
                 LU_Around = c("Pasture","Perennial","Perennial"),
                 Type = c("Burial Mound","Burial Mound","Burial Mound"),
                 Source = c("Excavation","Excavation","Excavation"))
extrasf <- st_as_sf(extras, coords =c("Easting", "Northing"), crs = 32635)

```

```{r BA-combo, include = FALSE}
# Combine both BA datasets and spatialize
library(plyr)
BApoints <- rbind.fill(BA, extrasf)
BAcoords <- rbind(st_coordinates(BA) ,st_coordinates(extrasf))
BApoints <- cbind(BApoints, BAcoords)
BApoints <- st_as_sf(BApoints, coords = c("X", "Y"), crs = 32635)
```

## Linestring creation

Let's first see if we an create linestrings to later signal which mounds can *see* each other. To do that we need two columns of coordinates: the origin and the target. Here we explore the visibility of all BA mound from BA mound (from Toshko)
```{r linestring}
# Test linestring creation from one origin and multiple target points. The coords object needs to be a matrix, but to cbind one to many coordinates the component columns need to be dataframes.

coords <- as.matrix(cbind(as.data.frame(st_coordinates(extrasf$geometry[2])),
                as.data.frame(st_coordinates(BApoints$geometry))))

lines_sm <-  st_sfc(
     lapply(1:nrow(coords),
           function(i){
             st_linestring(matrix(coords[i,],ncol=2,byrow=TRUE))
           }))

st_crs(lines_sm) <- st_crs(BApoints)


library(mapview)
mapview(lines_sm)+mapview(BApoints)

# https://stackoverflow.com/questions/65498300/how-to-efficiently-create-linestrings-from-points
# https://stackoverflow.com/questions/58150279/plotting-lines-between-two-sf-point-features-in-r
```
 
## Intervisibility

Intervisibility in human terms mean that two or more points in the landscape have a clear line of sight between them (given reasonable visibility). In computational terms, it means that a straight line connecting the z values of the A and B (the two end-points) does not intersect a polygon formed by terrain profile between them. If it doesn't, A can see B.
Let's test such line of sight by extracting data out of raster cells between our BA points.

https://stackoverflow.com/questions/21841387/r-code-that-evaluates-line-of-sight-los-between-two-lat-lon-points 

https://gis.stackexchange.com/questions/272122/performing-viewshed-analysis-in-r
```{r define-cansee-functions}
cansee <- function(r, xy1, xy2, h1=0, h2=0){
### can xy1 see xy2 on DEM r?
### Y_elev is a DEM in same x,y, z units
### xy1 and xy2 are 2-length vectors of x,y coords
### h1 and h2 are extra height offsets (ie. mound heights)
###  (eg top of mast, observer on a ladder etc)
    xyz = rasterprofile(r, xy1, xy2)
    np = nrow(xyz)-1
    h1 = xyz$z[1] + h1
    h2 = xyz$z[np] + h2
    hpath = h1 + (0:np)*(h2-h1)/np
    return(!any(hpath < xyz$z))
}



viewTo <- function(r, xy, xy2, h1=0, h2=0, progress="none"){
    ## xy2 is a matrix of x,y coords (not a data frame)
    require(dplyr)
    apply(xy2, 1, function(d){cansee(r,xy,d,h1,h2)}, .progress=progress)
}

viewTo <- function(r, xy, xy2, h1=0, h2=0){
    ## xy2 is a matrix of x,y coords (not a data frame)
    require(dplyr)
    apply(xy2, 1, function(d){cansee(r,xy,d,h1,h2)})
}

rasterprofile <- function(r, xy1, xy2){
### sample a raster along a straight line between two points
### try to match the sampling size to the raster resolution
    dx = sqrt( (xy1[1]-xy2[1])^2 + (xy1[2]-xy2[2])^2 )
    nsteps = 1 + round(dx/ min(res(r)))
    xc = xy1[1] + (0:nsteps) * (xy2[1]-xy1[1])/nsteps
    yc = xy1[2] + (0:nsteps) * (xy2[2]-xy1[2])/nsteps
    data.frame(x=xc, y=yc, z=r[cellFromXY(r,cbind(xc,yc))])
}
```
Let's test these with real data (BA points 8346, 8007 and Drazhevo)


```{r rasterprofile}
BApoints[c(3,6,10),]
m8007 <- c(st_coordinates(BApoints)[3,],h= BApoints$HeightMax[3])
m8346 <- c(st_coordinates(BApoints)[6,],h= BApoints$HeightMax[6])
mDrazh<- c(st_coordinates(BApoints)[10,],h= BApoints$HeightMax[10])

Y_elev

test_see <- cansee(Y_elev, m8007, m8346, h1 = m8007[3]  , h2 = m8346[3])
test_see

```
## Automating the visibility
```{r}
# origin point coordinates and height
Yam_mnds %>% 
  group_by(Type) %>% tally()

Yam_mnds$HeightMax[is.na(Yam_mnds$HeightMax)] <- 0


origin <- cbind(st_coordinates(Yam_mnds), h = Yam_mnds$HeightMax)
target <- origin 

class(origin)

result_table <- NULL
for (i in 1:nrow(origin)){
  for (e in 1:nrow(target)){
      result = cansee(Y_elev, origin[i,1:2], target[e,1:2], h1 = origin[i,3], h2 = target[e,3])
    result_row <- cbind(i,e, result) 
    result_table <- rbind(result_table, result_row)
  }
}
head(result_table)
tail(result_table)
saveRDS(result_table, "../output_data/Yam_mnd_intervis250k.rds")

```

Finishing the 1,000,000 calculations takes a while! (1 hr so far at 17.18 on Thursday night).
OK, some 24 hours later (Friday afternoon), 241 mounds (*1000) were completed. Clearly the tasks takes 4 days.

```{r}
# Check the results of the initial intervisibility run
result_table <- readRDS("../output_data/Yam_mnd_intervis250k.rds")
result_table <- as.tibble(result_table)

# 10 most commanding mounds! i is the index, n is the number of visible features from it
result_table %>% 
  group_by(i) %>% 
  tally(result) %>% 
  arrange(desc(n))
# maxima were i = 241, and e = 780 
result_table %>% 
  filter(i == 240) %>% 
  arrange(desc(e))


# calculations per hr
250000/3600

```
### Link results to actual mound locations

### Visualizing the terrain profile
```{r}
# Create a terrain profile and plot it to check potential intervisibility
testprofile <- rasterprofile(Y_elev,  m8007, m8346)
testprofile <- rasterprofile(Y_elev,  m8007, mDrazh)
testprofile <- rasterprofile(Y_elev,  mDrazh, m8346)

# Plot
plot(testprofile$z, type = "b"); 

# create the connecting line between origin and destination points
x = c(0,nrow(testprofile))
y = testprofile$z[c(1,nrow(testprofile))]

# Plot profile with start and end point elevation
plot(testprofile$z, type = "l", main = "Terrain profile between mounds 8346 and Drazhevo", 
     ylab = "Elevation(m)", xlab = "Mound distance (*30m)"); 
lines(x,y, col = "red", lwd = 2)

```

None of these mounds (Drazhevo and 8346) can see each other, because the one on top of the Zaichi Vrah is obscured by the hill and not aimed at the mounds to the SE. 
Let's test the intervisibility between these three and other mounds in the surroundings from the Yambol Mounds dataset. We will use the viewTo() function and the starting coordinates of the three mounds.

```{r view-to-8007, warning=FALSE}
# First we need to convert the Yambol mounds to a matrix of coordinates
coords <- data.frame(st_coordinates(Yam_mnds),h=Yam_mnds$HeightMax)
missingHeight <- which(is.na(coords$h))
mnd_missH <- Yam_mnds$TRAP[missingHeight]

coords <- coords %>% 
  filter(!is.na(h))
coords_m <- as.matrix(coords)

# viewTo() working with 3 points
testview <- viewTo(Y_elev, m8007, rbind(m8346,mDrazh))

# with all points (uncomment if you wish to regenerate!)
# testview <- viewTo(Y_elev, m8007, coords_m,m8007[3], coords_m[,3])
# length(testview)


# # See which ones can see 8007
# '%nin%' = Negate('%in%')
# viewTo_8007 <- Yam_mnds %>% 
#   filter(TRAP %nin% mnd_missH) %>%   # must first filter away missing heights as I did with coordinates
#   filter(testview) # use the logical vector to filter the Yam_mnds rows

# 289 mounds can see mound 8007
# saveRDS(viewTo_8007, "../output_data/MoundsVisfrom8007.rds")
viewTo_8007 <- readRDS("../output_data/MoundsVisfrom8007.rds")
```

View these 289 mounds visible from 8007
```{r plot-visible}
# Test linestring creation from one origin and multiple target points. The coords object needs to be a matrix, but to cbind one to many coordinates the component columns need to be dataframes.
coords <- as.matrix(cbind(rep(m8007[1], 289),rep(m8007[2], 289), st_coordinates(viewTo_8007)))

lines <-  st_sfc(
     lapply(1:nrow(coords),
           function(i){
             st_linestring(matrix(coords[i,],ncol=2,byrow=TRUE))
           }))

st_crs(lines) <- st_crs(BApoints)

library(mapview)
library(leafsync)
m1 <- mapview(lines) + mapview(viewTo_8007, cex = 4) + mapview(BApoints[3,], cex = 10)
m2 <- mapview(viewTo_8007, zcol = "prom250mbuff")
sync(m1,m2)
```

## Distance to zemlishte central settlement
AKB asks for distance and azimuth to the nearest village (in "zemlishte")

```{r load-distance}
# Zemlishte information sit in ElhStr_towns shapefile
village <- read_sf("../data/ElhStr_towns.shp")
data22 <- read_sf("../data/ElhStr_features.shp")
```

### Wrangle the features and nearest municipality centers (which need not be the geographically nearest places)
```{r}
# Subset the features 2022 data to only TRAP numbers and the name of the  zemlishte-defining village
features22 <- data22 %>%
  dplyr::select(TRAP,EKATTE, Name_en) #EKATTE and Name_en are the number and name of the zemlishte that we are interested in

# Create a simple feature of village points ordered by the TRAP numbers from the features 22 dataset,   
villages22 <- features22 %>% 
  st_drop_geometry() %>%  # dropping feature geometry
  left_join(village, by = "EKATTE") %>%  #joining village geometry to TRAP numbers
  dplyr::select(-Descr_bg, -Name_bg, -Suffix_bg, -Suffix_en)
villages22
```

### Calculate distances

between mounds and zemlishte using the st_distance() function, utilizing by_element argument, because I am only interested in pairwise distances 
```{r}
features22$distToTown <- st_distance(features22$geometry, villages22$geometry, by_element = TRUE)
ls('package:sf')
```

### Calculate azimuth

I use the nngeo package of Michael Dorman as it calculates "planar" /geometric azimuth (https://michaeldorman.github.io/nngeo/index.html).
sf library works in geometric /planar space and so 
The other geosphere::azimuth which calculates geographic azimuth in latlong coordinates.(https://stackoverflow.com/questions/51030060/in-geosphere-package-in-r-why-arent-bearings-in-0-360-degrees) 


```{r}
install.packages("remotes")
remotes::install_github("michaeldorman/nngeo")
features22$azimuth <- nngeo::st_azimuth(features22$geometry, villages22$geometry)
```

### Export for Toshko
```{r}
# All 479 features from the Elhovo and Straldzha municipality (2010-2022)
write.csv(features22 %>% st_drop_geometry(), "../output_data/features22forToshko.csv")

# Only 238 features collected in 2022 in the Elhovo and Straldzha municipality
library(mapview)
features22 %>% 
  filter(TRAP %in%mnd22$TRAP) %>% 
  mapview()

# Only 311 features collected in 2010-2022 in the Elhovo municipality
mun <- read_sf("../data/Yam_Municipalities.shp")

library(sf)
features22$geometry %>% 
  sf::st_intersection(mun[4,]) %>% 
  mapview()
```

