---
title: "Bunkers"
author: "Adela Sobotkova"
date: "2022-11-12"
output: html_document
---
```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


```{r libraries}
library(tidyverse)
library(sf)
library(raster)
```

## Load 2009-2022 mound data and 2018 - 2022 military feature data
```{r}
getwd()
Yam_mnds <- readRDS("../output_data/mnds_aggr.rds")
Yam_mil <- read_csv("../output_data/MIlitaryFeatures-OR12Dec22.csv")
```

### Explore the data a bit
```{r}
tail(Yam_mil)
Yam_mil <- Yam_mil[-79,]

unique(Yam_mil$FacingDirection)
unique(Yam_mil$CentralFeature)
#Yam_mil$feature.facing <- factor(Yam_mil$feature.facing)

Yam_mil %>% 
  group_by(CentralFeature) %>% 
  count() %>% 
  arrange(desc(n))
```

## Load elevation raster and Yambol region boundary
```{r get-elev}
Y_region <- st_read("../data/YamRegion.shp")

Y_elev <- raster("../output_data/large/Yelev32635.tif")
plot(Y_elev, main = " Yambol region with burial mound distribution");plot(Yam_mnds$geometry, cex = sqrt(as.numeric(Yam_mnds$HeightMax)), add= T)
```


## Plot
```{r}
Yam_bunkers <- Yam_mnds %>% 
  dplyr::filter(TRAP %in% Yam_mil$MoundID) %>% 
  full_join(Yam_mil, by = c("TRAP"="MoundID"))


library(mapview)
Yam_bunkers$CentralFeature <- factor(Yam_bunkers$CentralFeature
)
levels(Yam_bunkers$CentralFeature) <- c("Bunker", "Infantry trench","Tank emplacement","Vehicle emplacement")

mapview(Yam_mnds, cex = 1) + mapview(Yam_bunkers, zcol = "CentralFeature")
```
## Great, let's get really plotting

circle = bunkers, 19, 20, 21 - color fill
square = trench, 7, 22
u/horseshoe = emplacement, 18, 25
mund triangle - 2,17,24

Maps: 
1: Yambol with all mounds with an inset of BG (no topography, minimalist graphics)
2: Border region with mounds and military installations visually differentiated
3: Typology of mil.installations
4: Sitovo site with different installations?

## Yambol with an inset of BG - Leaflet

```{r}
library(leaflet)
library(leafem)
leaflet() %>% 
  addProviderTiles("Esri.WorldImagery") %>% 
  addCircleMarkers(data = st_transform(Yam_bunkers, 4326)) %>% 
  addMiniMap(toggleDisplay = TRUE,
             position = "bottomright")

```

## Bulgaria inset preparation
```{r}
BG <- getData('GADM', country = 'BG', level = 0)
BG <- st_transform(st_as_sf(BG), 32635)

#BGregion <- getData('GADM', country = 'BG', level = 2)
plot(BG$geometry, col = "lightgrey", border = "grey"); plot(Y_region$geometry,col = "darkgrey", border= "darkgrey", add = T)
```

Prepare to print with the inset
```{r}
library(tmaptools)
library(grid)

# change back to the plotting mode
#tmap_mode("plot")

# pdf("figures/Figure01.pdf", width = 5)
# PPAP
overview2

# print insets
print(greece, vp=viewport(x= 0.17, y= 0.21, width= 0.3, height= 0.3))

#dev.off()
```

Tmap differentiating Yambol mounds/features and military installations


```{r}
library(tmap)
tm_shape(Yam_mnds, bbox = st_bbox(Y_region))+
  tm_symbols(shape = 2, col="#666666", size = 0.02) +
tm_shape(Y_region) +
 tm_borders(col = "grey", 
              lwd = 2)+
tm_shape(Yam_bunkers, bbox = st_bbox(Y_region)) + 
  # tm_polygons(col="cornsilk2", border.alpha = 0)+
  # tm_shape(features) + 
  tm_symbols(shape = 2, col="red", size = 0.1) +  #6 and 26 are downward triangles
  tm_compass(type = "arrow", 
             position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"),
               breaks = c(0,5,10), 
               text.size = 0.75) + 
  #tm_credits("Sobotkova 2020", position = c("right", "bottom"), size = 0.8) +
  #tm_layout(title = "Registered standing features") 
  tm_layout(title = "A")

  
  tm_layout(main.title = "Yambol and Bulgaria",
            legend.outside = FALSE)
```
Local typology
```{r}
library(tmap)
tm_shape(Yam_bunkers, bbox = st_bbox(Yam_bunkers)) + 
  tm_dots(
          # col = "CentralFeature", 
          col = "grey",
           shape = "CentralFeature", #6 and 26 are downward triangles
           title = "Registered military features",
          alpha= 0.6,
             size = 0.3) +  
tm_shape(Y_region) +
 tm_borders(col = "grey",
              lwd = 2)+
  tm_compass(type = "arrow", 
             position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"),
               breaks = c(0,5,10), 
               text.size = 0.75)  +
  tm_layout(title = "B", legend.show = FALSE)


```
Try facets
```{r}
tm_shape(Yam_bunkers) + 
  tm_dots()+
  tm_facets(by = "CentralFeature",free.coords = FALSE)+
  tm_shape(Y_region) +
 tm_borders(col = "grey",
              lwd = 2) +
  tm_style("cobalt")

```

