---
title: "Excavated Mounds in the AOR"
author: "Adela Sobotkova"
date: "2022-12-14"
output: html_document
---

This rmarkdown is not usable yet, just dumping preliminary AOR loading code chunks. To be elaborated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## # Excavated mounds from Yambol based on AORs

```{r load-AOR-google, eval = FALSE}
library(googlesheets4)
gs4_deauth()


general <- read_sheet("https://docs.google.com/spreadsheets/d/1l2glM1X8eiWFArunnnTG6iGdWUvcd7LVBYiFcbjBDJQ/edit#gid=1761342759&fvid=1870098839")
mound <- read_sheet("https://docs.google.com/spreadsheets/d/1l2glM1X8eiWFArunnnTG6iGdWUvcd7LVBYiFcbjBDJQ/edit#gid=1172661582", range = "MoundAttributes")
burial <- read_sheet("https://docs.google.com/spreadsheets/d/1l2glM1X8eiWFArunnnTG6iGdWUvcd7LVBYiFcbjBDJQ/edit#gid=1172661582", range = "BurialAttributes")

# here I can join them all and play.
```
Alternatively, and better perhaps, I can load the merged data in json format from \url{https://sciencedata.dk/index.php/apps/files/?dir=%2FSDAM_root%2FSDAM_data%2Fmounds&id=7813458&owner=648597%40au.dk} sciencedata.dk, unlist and work with
These Izvestia_df.json and AOR_df.json have been downloaded in data folder
Some guidance on unnesting:
https://stackoverflow.com/questions/38860380/unnesting-a-list-of-lists-in-a-data-frame-column

```{r eval = FALSE}
library(jsonlite)

json <- read_json("../data/AOR_df_2020-06-24.json", simplifyVector = TRUE)

json <- json %>% as_tibble()

# test
j <- json %>% 
  select(MoundID, Municipality,Region, Lat, Long, `Error radius(m)`, 
    LU_Around, MoundCover, Geomorphology, Prominence, MoundName, GT, RT,Condition, 
    `Source of Impact`, FirstStartDate, 
   MaxEndDate, FirstEnclosureType, 
    ) %>% 
  unnest(cols = c(MoundID, Municipality,Region, Lat, Long, `Error radius(m)`, 
    LU_Around, MoundCover, Geomorphology, Prominence, MoundName,GT, RT,Condition, 
    `Source of Impact`,  FirstStartDate, 
   MaxEndDate, FirstEnclosureType, 
     ))
j
# numbers in columns seem to be a problem:

BApotential <- j %>% 
  filter(Region == "Yambol") %>%
  filter(FirstStartDate < -2000)
  
BApotential


```
